---
title: "Non standard evaluation"
subtitle: tidyeval
author: "A. Ginolhac, E. Koncina, R. Krause"
date: "4 May 2017"
output:
  iosp::ioslides_plus:
    box_colours:
      orange: "darkorange"
      slate: "slateblue2"
      yellow: ["yellow", "gold2"]
      yellow2: ["lightgoldenrod1", "goldenrod2"] # from Eric
      grgold: ["springgreen4", header_text: "#FFD700"]
---

```{r setup, include = FALSE}
library(tidyverse)
ggplot2::theme_set(ggplot2::theme_bw(18))
```


## lazy evaluation

the **tidyverse** makes your life easy by evaluating columns in `df`/`tibble` context.

### R base, must refers to `mtcars` {.box-6 .stretch}

```{r}
plot(mtcars$wt, mtcars$mpg)
```

### ggplot2 does the job {.box-6}

```{r}
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point()
```

## non standard evaluation is hard {.vs2}

### evaluation in base, problem {.box-6 .stretch .build .bg-red .compact-output}

`wt` is unknow in the global environment

```{r, error=TRUE}
wt
```

### evaluation in base, solution {.box-6 .stretch .build .bg-yellow2 .compact-output}

combine `substitute` and `eval` in _df_ context

```{r}

eval(substitute(wt), mtcars)
```

### evaluation in tidyverse {.box-12 .build .bg-grgold .compact-output}


```{r, row = TRUE}
mtcars %>%
  select(wt) %>%
  head()
```

### Drawback {.box-12 .build .bg-yellow2 }

- Make your life easy when performing **interactive** analysis 
- Hard when doing programming, as soon as you need **iteration** and hence **function**

# ggplot2

## The issue | ggplot2

When dealing with **global** variables, fine. But **local** ones in _function_...

### in global env {.box-6 .bg-cobalt .build}

```{r}
mtcars %>%
  ggplot(aes(x = mpg)) +
  geom_histogram(bins = 25)
```


### in function {.box-6 .bg-red .build}

```{r, error=TRUE}
mtcars_dens <- function(df, col) {
  df %>%
    ggplot(aes(x = col)) +
    geom_histogram(bins = 25)
}
mtcars_dens(df = mtcars, col = "mpg")
```

## Solution: `aes_string()` | only for ggplot2 {.vs1}

### {.box-12}

```{r, row=TRUE, fig.height=3}
mtcars_dens <- function(df, col) {
  df %>%
    ggplot(aes_string(x = col)) +
    geom_histogram(bins = 25)
}
mtcars_dens(df = mtcars, col = "mpg")
```


### {.box-12 .build}

```{r, row=TRUE, fig.height=3}
mtcars_dens(df = mtcars, col = "hp")
```


# dplyr < 0.6


## before upcoming 0.6

### in global env {.box-6 .bg-cobalt .build}

```{r}
mtcars %>%
  filter(hp > 250)
```


### in function {.box-6 .bg-red .build}

```{r, error=TRUE}
mtcars_filter <- function(df, col) {
  df %>%
    filter(col > 250)
}
mtcars_filter(df = mtcars, col = "hp")
```


## Solution: underscore versions | dplyr < 0.6 {.vs2}

### {.box-12}

```{r, row=c(8, 4)}
mtcars_filter <- function(df, col) {
  filter_call <- lazyeval::interp(~ var > 250, var = as.name(col))
  df %>%
    filter_(.dots = filter_call)
}
mtcars_filter(df = mtcars, col = "hp")
```

### Obscure {.box-6 .bg-orange .build}

Ugly, complex, hard but mandatory

### Proceed anyway {.box-6 .bg-orange .build .stretch}

With dynamic threshold

%end%

### {.box-12 .build}

```{r, row=c(8, 4)}
mtcars_filter <- function(df, col, threshold) {
  filter_call <- lazyeval::interp(~ var > thr,
                                  var = as.name(col),
                                  thr = quote(threshold))
  df %>%
    filter_(.dots = filter_call)
}
mtcars_filter(df = mtcars, col = "hp", threshold = 265)
```

# dplyr >= 0.6

## tidyeval | dplyr 0.6, upcoming! {.vs2}

described in this [vignette](http://dplyr.tidyverse.org/articles/programming.html)

### classic quoting {.box-6 .bg-blue .build}

variables as strings

```{r}
fun1 <- function(name) {
  paste0("name is, ", name)
}
fun1("stringent")
```


### introducing `glue`{} {.box-6 .bg-yellow2 .build .stretch}

```{r}
fun2 <- function(name) {
  glue::glue("name is {name}")
}
fun2("stringent")
```


## Solution with tidyeval {.vs1 .compact-output}
 
 
currently [discussed](https://github.com/tidyverse/dplyr/issues/2726)

We can pass on `hp` as a promise (it is unknow in **Global Env**), evaluation is delayed and `enquo()` uses 
some _dark magic_ 


```{r, error=TRUE}
 
mtcars_filter <- function(df, column) {

  col <- enquo(column)
  message(col)
  df %>%
    select(!! col)
}
mtcars_filter(df = mtcars, column = hp)
```


## Expressions


### `quo`inside function {.box-6} 

```{r, error=TRUE}
mtcars_filter <- function(df, column, threshold) {

  var <- quo(hp  > !! threshold)
  message(var)
  df %>%
    filter(!! var)
}
mtcars_filter(df = mtcars, hp, threshold = 300)
```

### `quo`ins function' call {.box-6} 

```{r}
mtcars_filter2 <- function(df, expr) {
  var <- enquo(expr)
  message(var)
  df %>%
    filter(!! var)
}
mtcars_filter2(df = mtcars, cyl > 6)
mtcars_filter2(df = mtcars, hp > 250)
mtcars_filter2(df = mtcars, am == 0)
```



## different output / input variables

- to turn a **quosure** into a **name** that could be pasted
- Must uses a specific assignment `:=`

```{r}
my_mutate <- function(df, out_name, expr) {
  expr <- enquo(expr)
  out <- enquo(out_name)
  g_name <- paste0("mean_", quo_name(out))

  summarise(df, 
    !!g_name := mean(!!expr)
  )
}

my_mutate(mtcars, out_name = gear_type, am)
```



## Pass arguments as promises, not strings {.vs3}


> I can categorically say if you're pasting strings to program with dplyr, there is always better way. | [Hadley](https://github.com/tidyverse/dplyr/issues/2663#issuecomment-294206288)
