---
title: "Data processing with R tidyverse"
subtitle: "Nothing on stringr, really"
author: "G. Androsova, R. Krause"
date: "5 May 2017"
output:
  iosp::ioslides_plus:
    footer: "[R tidyverse workshop](https://lsru.github.io/tv_course/)"
    box_colours:
      bg-yellow: ["lightgoldenrod1", "goldenrod2"]
      bg-grayblue: ["#bfccd7"]
      bg-bluewhite: ["white", "#005c99"]
      bg-greenwhite: ["white", "#1f7a1f"]
      bg-elixir: [  "white", "#023452", "#ED7520"]
      yellow2: ["lightgoldenrod1", "goldenrod2"]
---

## Learning objectives


### A first glance at regular expressions {.box-8 .offset-1 .bg-bluewhite}
 * Matching substituting of character strings 
 * `s/^lecture([0-9]{1,2}).*[^_].Rmd$/\1.Rmd/g`
 * Ugly, unreadable, terrifying expression 
 * See [R for data science](http://r4ds.had.co.nz/strings.html#matching-patterns-with-regular-expressions)
 


## Motivation
### What if we want to match 
- any letter followed by 'n'?
- any vowel followed by 'n'?
- two letters followed by 'n'?
- any number of letters followed by 'n'? 

### Regular expressions!
- allow us to match much more complicated patterns 
- build patterns from a simple vocabulary and grammar 

## Finite state automaton

### Most relevant consequence {.bg-bluewhite .box-6}
1. It will always return the
earliest (leftmost) match it
finds. 


2. Given a choice it always
favors match over a nonmatch

### Examples {.bg-bluewhite .box-6}
The **top**ic of the day is isotopes.


## Typical R functions using regex

### `grep, grepl` {.box-6 .bg-bluewhite .build}
Search for matches of a regular expression/pattern in a character vector; either return the indices into the character vector that match, the strings that happen to match, or a TRUE/FALSE vector indicating which elements match
```{r}
grep("c", "cat")
```

```{r}
grep("a", "cat", value = TRUE)
```

```{r}
grepl("at", "cat")
```

### `sub, gsub` {.box-6 .bg-greenwhite} 
Search a character vector for regular expression matches and replace that 
match with another string
```{r}
sub("c", "h", "cat")
```

```{r}
sub("c", "kl", "cataract")
```

```{r}
gsub("c", "n", "cataract")
```

## Meta-characters | Not representing matched characters
### `.` (period) represents any character except empty string ''

```{r}
vec <- c('ct', 'cat', 'cart', 'dog', 'rat', 'carert', 'bet')
grep(".at", vec)
grep("..t", vec)
```
`+` represents one or more occurrences

```{r}
grep('c.+t', vec)
```

## Meta characters
\* represents zero or more occurrences

```{r}
vec 
grep('c.*t', vec)
```
Group terms with parentheses '(' and ')'

```{r}
grep('c(.r)+t', vec)
grep('c(.r)*t', vec)
```

## Quantifying number of matches

### Applies to preceeding of group {.box-6 .bg-yellow2 } 
 * `?` The preceding item is optional and will be matched at most once.
 * `*` The preceding item will be matched zero or more times.
 *`+` The preceding item will be matched one or more times.
 * `{n}` The preceding item is matched exactly 'n' times.
 * `{n,}` The preceding item is matched 'n' or more times.
 * `{n,m}` The preceding item is matched at least 'n' times, but not more than 'm' times.

## Match positions

### Other useful ones include {.box-6 .bg-bluewhite}
 * `^` Start of string
 * `$` End of string

### Other useful ones include {.box-6 .bg-cobalt}

```{r}
vec
grep('r.$', vec)
```


```{r}
grep('^c', vec)
```

## Meta characters
 * `|` ( logical OR )
```{r}
vec
grep('(c.t)|(c.rt)', vec)
```

## Character classes
 * **[a-z]** lowercase letters
 * **[a-zA-Z]** any letter
 * **[0-9]** any number'
 * **[aeiou]** any vowel
 * **[0-7ivx]** any of 0 to 7, i, v, and x

Inside a character class **??** means anything except the following characters. E.g.

**[??0-9]** anything except a digit

## Matching metacharacters
We saw a bunch of special characters . + * ] [ $ What if we want to match them?
```{r}
vec2 = c("ct", "cat", "caat", "caart", "caaat", "caaraat", "c.t")
grep('c.t', vec2)
```
Escape them with \ or in R's case, two \\

```{r}
#grep('c\.t', vec) will not work, because R thinks \. is a special character like \n

#Use two \'s
grep('c\\.t', vec2)
```

## Matching metacharacters

To match a \, our pattern must represent \\

Our string must contain 4 backslashes!
```{r}
vec = c("a\\backslash", "nobackslash")
#grep('\\', vec) gives error
grep('\\\\', vec)
```

## Search and replace
### The **sub** function allows search and replacement {.box-12 .bg-blue}
```{r}
vec2
sub('a+', 'a', vec2)
```
**sub** replaces only first match, **gsub** replaces all

Use the backreferences `\1, \2` etc to refer to first, second group, etc.
```{r}
gsub('(a+)r(a+)', 'b\\1brc\\2c', vec2)
```

## Search and replace
### Use `\U, \L, \E` to make following backreferences upper or lower case or leave unchanged {.box-12 .bg-cobalt}
```{r}
vec2
gsub('(a+)r(a+)', '\\U\\1r\\2', vec2)
gsub('(a+)r(a+)', '\\U\\1r\\E\\2', vec2)
```

## Summary
**.** stands for any character.

**[ABC]** means A,B or C.

**[A-Z]** means any upper letter between A and Z.

**[0-9]** means any digit between 0 and 9.

List of metacharacters '**$ * + . ? [ ] ^ { } | ( ) \'**. If you need to use one of those characters, precede them with a doubled backslash.

## Extended list of regular expressions

Requires `perl=TRUE` flag

### Readable short cuts

**[:digit:]** Digits: '0 1 2 3 4 5 6 7 8 9'.

**[:alpha:]** Alphabetic characters: '[:lower:]' and '[:upper:]'.

**[:upper:]** Upper-case letters.

**[:lower:]** Lower-case letters.

Note that the set of alphabetic characters includes accents such as ร, รง or รถ which are very common is some languages. Therefore, it is more general than **[A-Za-z]** which ascii characters only.

## Extended list of regular expressions
### For other characters {.box-10 .bg-blue}

**[:punct:]** Punctuation characters: '! " # $ % & ' ( ) * + , - . / : ; < = > ? @ [ \ ] ^ _ ` { | } ~'.

**[:space:]** Space characters: tab, newline, vertical tab, form feed, carriage return, and space.

**[:blank:]** Blank characters: space and tab.

## Extended list of regular expressions
### For combination of other classes {.box-10 .bg-blue}

**[:alnum:]** Alphanumeric characters: '[:alpha:]' and '[:digit:]'.

**[:graph:]** Graphical characters: '[:alnum:]' and '[:punct:]'.

**[:print:]** Printable characters: '[:alnum:]', '[:punct:]' and space.

**[:xdigit:]** Hexadecimal digits: '0 1 2 3 4 5 6 7 8 9 A B C D E F a b c d e f'.



## Concatenating strings
### Base R 
 * `paste()` concatenates strings.
 * `paste0()`
 * `cat()` prints and concatenates strings
```{r}
paste("toto","tata",sep=' ')
paste("toto","tata",sep=",")
x <- c("a","b","c")
paste(x,collapse="-")
cat(c("a","b","c"), sep = "+")
```

## Splitting a string

**strsplit( )** : Split the elements of a character vector 'x' into substrings according to the matches to substring 'split' within them.
**cat()** prints and concatenates strings
```{r}
strsplit("a.b.c", "\\.")
unlist(strsplit("a.b.c", "\\."))
```

## Mastering regular expressions

### Working with regular   {.box-6 .bg-red}
 1. Start simply and expand
 2. Think about the most comprehensive -- shortest -- expression
 3. Think about negation rather than inclusion of all possibilities






