---
title: "Non standard evaluation"
subtitle: tidyeval
author: "A. Ginolhac, E. Koncina, R. Krause"
date: "4 May 2017"
output:
  iosp::ioslides_plus:
    box_colours:
      orange: "darkorange"
      slate: "slateblue2"
      yellow: ["yellow", "gold2"]
      yellow2: ["lightgoldenrod1", "goldenrod2"] # from Eric
      grgold: ["springgreen4", header_text: "#FFD700"]
---

```{r setup, include = FALSE}
library(tidyverse)
ggplot2::theme_set(ggplot2::theme_bw(18))
```

# ggplot2

## The issue | ggplot2

When dealing with **global** variables, fine. But **local** ones in _function_...

### in global env {.box-6 .bg-cobalt .build}

```{r}
mtcars %>%
  ggplot(aes(x = mpg)) +
  geom_histogram(bins = 25)
```


### in function {.box-6 .bg-red .build}

```{r, error=TRUE}
mtcars_dens <- function(df, col) {
  df %>%
    ggplot(aes(x = col)) +
    geom_histogram(bins = 25)
}
mtcars_dens(df = mtcars, col = "mpg")
```

## Solution: `aes_string()` | only for ggplot2 {.vs1}

### {.box-12}

```{r, row=TRUE, fig.height=3}
mtcars_dens <- function(df, col) {
  df %>%
    ggplot(aes_string(x = col)) +
    geom_histogram(bins = 25)
}
mtcars_dens(df = mtcars, col = "mpg")
```


### {.box-12 .build}

```{r, row=TRUE, fig.height=3}
mtcars_dens(df = mtcars, col = "hp")
```


# dplyr < 0.6


## before upcoming 0.6

### in global env {.box-6 .bg-cobalt .build}

```{r}
mtcars %>%
  filter(hp > 250)
```


### in function {.box-6 .bg-red .build}

```{r, error=TRUE}
mtcars_filter <- function(df, col) {
  df %>%
    filter(col > 250)
}
mtcars_filter(df = mtcars, col = "hp")
```


## Solution: underscore versions | dplyr < 0.6 {.vs2}

### {.box-12}

```{r, row=c(8, 4)}
mtcars_filter <- function(df, col) {
  filter_call <- lazyeval::interp(~ var > 250, var = as.name(col))
  df %>%
    filter_(.dots = filter_call)
}
mtcars_filter(df = mtcars, col = "hp")
```

### Obscure {.box-6 .bg-orange .build}

Ugly, complex, hard but mandatory

### Proceed anyway {.box-6 .bg-orange .build .stretch}

With dynamic threshold

%end%

### {.box-12 .build}

```{r, row=c(8, 4)}
mtcars_filter <- function(df, col, threshold) {
  filter_call <- lazyeval::interp(~ var > thr,
                                  var = as.name(col),
                                  thr = quote(threshold))
  df %>%
    filter_(.dots = filter_call)
}
mtcars_filter(df = mtcars, col = "hp", threshold = 265)
```

# dplyr >= 0.6

## tidyeval | dplyr 0.6, upcoming! {.vs2}

described in this [vignette](http://dplyr.tidyverse.org/articles/programming.html)

### classic quoting {.box-6 .bg-blue .build}

variables as strings

```{r}
fun1 <- function(name) {
  paste0("name is, ", name)
}
fun1("stringent")
```


### introducing `glue` {.box-6 .bg-yellow2 .build .stretch}

```{r}
fun2 <- function(name) {
  glue::glue("name is {name}")
}
fun2("stringent")
```


 ## Solution with tidyeval
 
 
```{r}
 
mtcars_filter <- function(df, col, threshold) {

  column <- quo(col > threshold)
  print(enquo(column))
  thr <- quo(threshold)
  df %>%
    filter(!!column)
}
mtcars_filter(df = mtcars, hp, threshold = 265)
```
