---
title: '`purrr` practical'
author: "Eric Koncina"
date: "4 May 2017"
output:
  unilur::tutorial_pdf: default
  unilur::tutorial_pdf_solution: default
---

```{r setup, include = FALSE}
library("tidyverse")
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

```{css}
h1 {
  font-size: 25px;
}
h2 {
  font-size: 20px;
}
h3 {
  font-size: 18px;
}
.bonus {
  color: red;
  text-decoration: underline;
}
.solution > h2 {
  font-size: 30px;
}
```

# 1. Loading data

Download a set of qPCR experiments results [here](https://lsru.github.io/tv_course/data/pcr.zip). The file is a zip archive which you should uncompress into a subfolder of your project.

For example, if you downloaded the file into your project folder you can run the following command:

```{r, eval = FALSE}
unzip("pcr.zip", exdir = "data/pcr") # will create a data/pcr subfolder and extract the files
```

```{r, echo = FALSE}
unzip(file.path("data", "pcr.zip"), exdir = "data/pcr")
```

These qPCR results were obtained from 2 different samples and were replicated 5 times. They are stored in 10 different files. A filename looks like "mlc1_1.csv" where the first number is the sample id and the second the replicate id.

## Read in the qPCR results

### Identify the file format

1. Using a text editor or RStudio, try to identify how the flat file has been encoded.
2. Try to import the file `mlc1_1.csv` using the `read_delim()` function.

```{r, solution = TRUE}
read_delim(file.path("data", "pcr", "mlc1_1.csv"), delim = " ")
```

### Import multiple files

3. Create a vector named `pcr_files` containing the path to all 10 data files using the `list.files()` function and adjust the `full.names` argument accordingly (if required you might get some help using `?list.files`).
```{r, solution = TRUE}
pcr_files <- list.files(file.path("data", "pcr"), full.names = TRUE)
```
4. Now use `map()` from the `purrr` package to import **all** files.
    + What is the type of the output (you might want to use `glimpse()`)
    + Are you able to identify from which file each element has been imported ?
```{r, solution = TRUE}
options(tibble.print_min = 3)
list.files(file.path("data", "pcr"), full.names = TRUE) %>%
  map(read_delim, delim = " ", col_types = cols()) %>% 
  str(max.level = 1)
# map() creates a list
# We are not able to tell which list element comes from which file
```
5. `map()` will name each output element according to the names found in the input vector. Use `set_names()` to keep this information.
```{r, solution = TRUE}
list.files(file.path("data", "pcr"), full.names = TRUE) %>%
  set_names() %>%
  map(read_delim, delim = " ") %>% 
  names()
```
- <span class = "bonus">(Supplementary question)</span> remove the path and extension from the filename using `basename()` and `tools::file_path_sans_ext()`.

```{r, solution = TRUE}
list.files(file.path("data", "pcr"), full.names = TRUE) %>%
  set_names(nm = (basename(.) %>% tools::file_path_sans_ext())) %>%
  map(read_delim, delim = " ") %>% 
  names()
```


6. Getting a single `tibble` out of all files would be much handier. Instead of further transforming this output we will use another member of the `map()` family of functions that will immediately create the desired output. Replace your call to `map()` in your previous code by the alternative member of the family to get a data frame directly.
```{r, message = FALSE, solution = TRUE}
list.files(file.path("data", "pcr"), full.names = TRUE) %>%
  set_names(nm = (basename(.) %>% tools::file_path_sans_ext())) %>%
  map_df(read_delim, delim = " ")
```
7. Are you still able to identify the different samples and replicates? You probably need to adjust the appropriate argument in your mapping function (have a look at the help page).
```{r, solution = TRUE}
# We are missing the sample and replicate variables (i.e. the associated filename)
# We need to adjust the .id argument to get the name of each individual
# tibble in a new column.
list.files(file.path("data", "pcr"), full.names = TRUE) %>%
  set_names(nm = (basename(.) %>% tools::file_path_sans_ext())) %>%
  map_df(read_delim, delim = " ", .id = "filename")
```

### Rearrange the data and save multiple files

Now that we were able to read in our data into a single data frame we would like to group together the measures for each individual gene and store them as separate `.csv` files.

8. first nest the data to let the measures associated to each gene appear in their own tibble.
9. create a folder to store the files
```{r, boxtitle = "Tip", box = "lightblue"}
# First we create a new folder inside the data folder to store the output files
# You can create such a folder relative to your project or Rmarkdown folder using the following command:
dir.create(file.path("data", "by_gene"), showWarnings = FALSE) 
# You can set `showWarnings = FALSE` to avoid a warning each time you knit
# or execute the chunk telling that the folder already exists 
```

```{asis, boxtitle = "Tip", box = "lightblue"}
In this tutorial we stored our input files in the `data/pcr` subfolder relative to the projects path.
To build up a platform independent path (windows uses `\` while linux and MAC OS are using `/`) we can use the function `file.path()` which will choose the appropriate path separator.
```
10. create a new column containing the path to the target name
    + first create a filename using `paste0`
    + then combine the folder path with the filename using `file.path()`.
11. now we should use a function able to write `.csv` files and map it to the appropriate vector(s). If `read_csv()` is able to read `.csv` files, which would be the `readr` function able to write such a file?
    + Is writing a file to a disk a side-effect? What would be the appropriate `purrr` function?

```{asis, solution = TRUE}
- Writing a file to a disk is considered to be a side-effect. We are indead **not interested in changing our data** and should use `walk` instead of `map`.
- We will use `write_csv()` with two arguments: 1) the data and 2) the file path.
- Here we will try two different approaches (and a third one which is actually only working with the developement version of `purrr` available on [github](https://github.com/tidyverse/purrr))
    + **First method** will use `walk2()` with two input lists (data and filename). We will have to put it inside a `mutate()` function within our workflow. This might not be the nicest place for this function as we don't want to mutate anything when calling `walk2()`. `walk2()` returns the first argument (data) leaving it unchanged and we will assign this output to the data column (this will overwrite it with itself leaving it unchanged).
    + **Second method** will use `walk()` but outside a mutate. We will use the **transposed** tibble as the input list (will map a function on each row instead of column) and declare an anonymous function using `write_csv()`.
    + **Third method _not working with the actual purrr found on CRAN_** will use the new function `iwalk()`. `iwalk()` is a shortcut to `walk2(.x, names(.x), .f)` to use the index.
```


```{r, solution = TRUE, eval = FALSE}
# Method 1: using walk2 but inside a mutate call
list.files(file.path("data", "pcr"), full.names = TRUE) %>%
  set_names(nm = (basename(.) %>% tools::file_path_sans_ext())) %>%
  map_df(read_delim, delim = " ", .id = "filename") %>%
  group_by(gene) %>%
  nest() %>%
  mutate(file_out = paste0(gene, ".csv"),
         file_out_path = file.path("data", "by_gene", file_out),
         data = walk2(data, file_out_path, write_csv))
```

```{r, solution = TRUE, eval = FALSE}
# Method 2: using walk on the transposed tibble with an anonymous function
list.files(file.path("data", "pcr"), full.names = TRUE) %>%
  set_names(nm = (basename(.) %>% tools::file_path_sans_ext())) %>%
  map_df(read_delim, delim = " ", .id = "filename") %>%
  group_by(gene) %>%
  nest() %>%
  mutate(file_out = paste0(gene, ".csv"),
         file_out_path = file.path("data", "by_gene", file_out)) %>%
  transpose() %>%
  walk(~write_csv(.$data, .$file_out_path))
```

```{r, echo = FALSE, eval = FALSE}
# To test the new version of purrr
# First install it in an alternative lib path
#dir.create("test-library")
#withr::with_libpaths(new = "test-library", devtools::install_github("tidyverse/purrr"))
unloadNamespace("modelr")
unloadNamespace("purrr")
library("purrr", lib.loc="test-library")
```


```{r, solution = TRUE, eval = FALSE}
# Method 3: using `iwalk()` (new purrr function)
list.files(file.path("data", "pcr"), full.names = TRUE) %>%
  set_names(nm = (basename(.) %>% tools::file_path_sans_ext())) %>%
  map_df(read_delim, delim = " ", .id = "filename", col_types = cols()) %>%
  group_by(gene) %>%
  nest() %>%
  mutate(file_out = paste0(gene, ".csv"),
         file_out_path = file.path("data", "by_gene", file_out)) %>%
  select(file_out_path, data) %>%
  deframe() %>% # generate a named list from a 2 column tibble
  iwalk(write_csv)
```

# 2. Loading data and handling untidy data

Download the excel file called [`sizes.xls`](https://lsru.github.io/tv_course/data/sizes.xls). The file contains measures (width, height and depth) of 5 different samples before and after a treatment. We would like to calculate the volume of the object before and after the treatment.

## Load the content of the excel file

First load the measures into a data frame using the powerful tools provided by the `purrr` package.

- What function would you like to repeat?
- How do you generate your starting vector?

```{r, solution = TRUE}
library(readxl)

xls_file <- "data/sizes.xls"

excel_sheets(xls_file) %>%
  set_names() %>% 
  map_df(read_excel, path = xls_file, .id = "treatment_onset", range = "B3:C9")
```

## Calculate the volume

It turns out that the size ($width \times height \times depth$) was not entered in a tidy form. Extract the different values in order to calculate the volume.

```{asis, boxtitle = "Warning", box = "orange"}
We already saw the `tidyr::separate()` function which might help you to get what you want.
Here we would like to stick to a `purrr` approach. Think about how many elements you would like to provide and how many elements you would like to get.
```

```{asis, boxtitle = "Tip", box = "lightblue"}
- You might want to use `base::strsplit()` or `stringr::str_split()` 
```

```{r, solution = TRUE}
# The tidyr / dplyr way but it's not a purrr approach
excel_sheets(xls_file) %>%
  set_names() %>% 
  map_df(read_excel, path = xls_file, .id = "treatment_onset", range = "B3:C9") %>% 
  separate(size, c("width", "height", "depth"), sep = "x") %>%
  mutate_at(c("width", "height", "depth"), parse_integer) %>%
  mutate(volume = width * height * depth)
```

```{r, solution = TRUE}
excel_sheets(xls_file) %>%
  set_names() %>% 
  map_df(read_excel, path = xls_file, .id = "treatment_onset", range = "B3:C9") %>% 
  mutate(tidy_size = stringr::str_split(size, "x") %>% map(parse_integer),
         volume = map_dbl(tidy_size, reduce, `*`))
```


```{r, echo = FALSE}
knitr::knit_exit()
```

### Tibbles are very flexible

To show you the power of tibbles and `purrr` we now would like to generate boxplots of the _Ct_ values for each type.

1. We would like to generate a distinct plot for the data from each file and store it directly in a tibble column.
    + build up an anonymous function which will create the plot
    + just store the output list `map()` returns as usual and have a look at the resulting tibble
    
```{r}
list.files("data/pcr", full.names = TRUE) %>%
  set_names() %>%
  map_df(read_delim, delim = " ", .id = "filename") %>%
  group_by(filename) %>%
  nest() %>%
  mutate(plot = map(data, ~ggplot(data = ., aes(x = type, y = Ct)) + geom_boxplot()))
```

2. Like in the previous question, we now would like to save each plot as an individual file.
    + create again a new subfolder
    + first build-up the path to the output file (original filename but with the `.pdf` extension)
    
```{r, echo = FALSE}
dir.create(file.path("data", "plots"), showWarnings = FALSE)
```

```{r, solution = TRUE}
list.files(file.path("data", "pcr"), full.names = TRUE) %>%
  set_names() %>%
  map_df(read_delim, delim = " ", .id = "filename") %>%
  group_by(filename) %>%
  nest() %>%
  mutate(plot_file = paste0(filename, ".pdf"),
         plot_file_path = file.path("data", "plots", plot_file),
         plot = map(data, ~ggplot(data = ., aes(x = type, y = Ct)) + geom_boxplot()),
         plot_file_path = walk2(plot_file_path, plot, ggsave, device = "pdf"))

list.files(file.path("data", "pcr"), full.names = TRUE) %>%
  map_df(read_delim, delim = " ", .id = "filename", col_types = cols()) %>%
  group_by(filename) %>%
  nest() %>%
  mutate(plot_file = paste0(filename, ".pdf"),
         plot_file_path = file.path("data", "plots", plot_file),
         plot = map(data, ~ggplot(data = ., aes(x = type, y = Ct)) + geom_boxplot())) %>%
  transpose() %>%
  walk(~ggsave(.$plot_file_path, .$plot, device = "pdf"))

# Or iwalk...
```


